The unphysicality of Hilbert spaces

### 2023-06-21 Articles/2023-HilbertQuantum/HilbertQuantum.tex
Cleaning up the appendix

We give the full mathematical detail in support of the argument laid out in the main paper

A vector space is a set where the elements can be summed and multiplied by a constant. In quantum mechanics, the set represents pure states and the linear combination represent superposition of states. One may already accept this as physical, as superpositions can be created in a lab when, for example, a particle goes through a beam splitter. However, there are two difficulties.

First of all, it is not clear whether superpositions are actual physical entities or whether they are just mathematical constructs. This is an area where different interpretations may differ, and we do not want our physical justification to depend on our interpretation. We want to decide whether the mathematical representation is physical regardless of the interpretation. Second, the actual states are not represented by vectors, but by rays. The superposition, then, is in terms of objects that are not themselves strictly physical. For these two reasons, we want to find another way an equivalent characterization of the linearity of the space that is fully physical.

Let us consider the space of density operators, which can be used to characterize both pure and mixed state. These objects are invariant under change of phase and are in one-to-one correspondence with the rays of the space, not with the vectors. The density operators have a linear structure that represent statistical mixtures. This is a necessary structure because preparing statistical mixtures is something we can always do physically, and therefore something a physically meaningful mathematical definition must provide. However, the linear structure of density operators is not the same as the linear structure of the Hilbert space.

When we write
$$c_1|\psi_1\> + c_2 |\psi_2\>$$
and 
$$p_1|\psi_1\>\<\psi_1| + p_2|\psi_2\>\<\psi_2|$$
the meaning of the $+$ symbols is different. The first operations is a complex linear combination of two pure states that returns another pure state, while the second is a real linear combination of two pure states that returns a mixed state. However, the two are so intertwined that one can be recovered by the other.

As we show in the appendix, a pure state $\psi$ is a superposition of two other pure states $\phi_1$ and $\phi_2$ if and only if there exists another pure states $\bar{\psi}$ such that we can mix $\psi$ and $\bar{\psi}$ to obtain a mixed state that can also be obtained by mixing $\phi_1$ and $\phi_2$. Geometrically, consider a Bloch sphere. If $\psi$ is a superposition of $\phi_1$ and $\phi_2$ then all three states are on the surface of the same sphere. The chord connecting $\phi_1$ and $\phi_2$ represents the space of all possible mixtures of $\phi_1$ and $\phi_2$. Now, take a point $\rho$ on that chord, and consider the line that passes through $\rho$ and $\psi$. The same line will traverse the sphere at another point, which we call $\bar{\psi}$. Since $\rho$ is on the chord that connects $\psi$ and  $\bar{\psi}$, it is a mixture of the two. Therefore the vector space structure can be recovered by knowing how pure state mix.

Note that the fact that the same statistical mixture can be decomposes in multiple ways is a characteristic feature of quantum mechanics, that is not present in classical mechanics. Using the same characterization, then, we consistently conclude that in classical mechanics there are no superpositions precisely because, given a pure state $\psi$, we can never find a corresponding state $\bar{\psi}$.

The argument, then, goes like this. On physical grounds, we must be able to construct mixture. On physical grounds, we must be able to tell whether the same mixture can be prepared in different ways. The vector space structure captures exactly which pairs of pure states produce indistinguishable mixtures. The vector space structure exactly captures that structure and therefore it is fully physically justified. \textbf{The vector space structure is physical.}

\subsection{Inner product}

The inner product defines a geometrical structure on top of the vector space. Namely, it gives an idea of size of each vector and of angle between vectors. In quantum mechanics, the square of the inner product
\begin{equation}
	p(\psi_1, \psi_2) = |\<\psi_1| \psi_2\>|^2.
\end{equation}
represent the probability of measuring one state having prepared another. One may already accept this as physical, as these probabilities define all observable features of quantum mechanics. There is, again, a difficulty.

Understanding the inner product as giving the probability of measurements also brings in one of the most controversial aspects of the theory: measurements. This is, again, a problem since we want our reasoning to be independent of interpretation. Another issue is that, technically, the inner product is linked to decomposition in linear combinations, to superposition, which are done at equal time. The probability above, instead, is a probability of transition.

Given that we have characterized superposition in terms of mixtures, we should see whether the inner product characterizes something fundamental about them, and indeed it does. Given two pure states, the entropy of the mixture is a strictly monotonic, and therefore invertible, function of the square of the inner product. The greatest entropy is given by mixing orthogonal states, which gradually decreases as the angle in the inner product space decreases. We can then understand the geometrical structure of the pure quantum states as an entropic structure, much like it happens for classical mechanics.\footnote{In classical mechanics, the Liouville measure returns the count of states within a region, which is used in the fundamental postulate of statistical mechanics. The symplectic form gives us the Liouville measure for each independent degree of freedom.}

The argument, then, goes like this. On physical grounds, for each statistical mixture, the thermodynamic entropy must be defined. Therefore it must be defined for equal mixtures of two pure states as well. The inner product exactly captures how that quantity combines and therefore it is fully physically justified. \textbf{The inner product structure is physical.}

@@

First we want to show that the vector space structure is not, by itself, physical. Since quantum states are represented by rays of a Hilbert space, and since rays can be defined simply on top of the vector space structure alone, one may think that the vector space structure, by itself, is physical. The issue is that, without an inner product, one cannot make physical sense of the coefficients one uses for linear combinations, therefore superpositions becomes ill-defined.

The following proposition tells us that there are infinitely many ways to decompose a vector representing a state in terms of vectors that represent two other states.
\begin{prop}\label{vector_insufficient}[The vector space structure is not enough to uniquely identify superpositions.] Let $\mathcal{H}$ be a complex vector space and let $\textbf{P}(\mathcal{H})$ be its projective space. Let $\psi, \phi_1, \phi_2 \in \mathcal{H}$ such that $\psi = c_1 \phi_1 + c_2 \phi_2$ for some $c_1, c_2 \in \mathbb{C}$. Let $\pj{\psi}, \pj{\phi_1}, \pj{\phi_2} \in \textbf{P}(\mathcal{H})$ be the corresponding rays in the projective space. Then for any $\hat{c}_1, \hat{c}_2 \in \mathbb{C}$ we can find $\hat{\phi}_1 \in \pj{\phi_1}$ and $\hat{\phi}_2 \in \pj{\phi_2}$ such that
$$\psi = \hat{c}_1 \hat{\phi}_1 + \hat{c}_2 \hat{\phi}_2.$$
\end{prop}

@@

The intuition that a state has a single linear decomposition in terms of other vectors requires, first of all, that the vectors are normalized. To identify the coefficient as probability amplitudes, that is that $\sum_i |c_i|^2 = 1$, further requires the decomposition to be done on orthogonal vectors.\footnote{TODO: mention Gram-Schmidt orthonormalization} That is, linear combinations require the inner product to be understood physically as superpositions. The vector space structure by itself is therefore insufficient to provide a clear connection to physics.

@@

First of all, it is not clear whether superpositions are actual physical entities or whether they are just mathematical constructs. This is an area where different interpretations may differ,\footnote{TODO: add citations} some arguing that they are just mathematical representations, and the physicality of a mathematical representation should not be interpretation dependent. Second, the Born rule, in its standard characterization, brings in one of the most controversial aspects of the theory, measurements, which suffers from an interpretation problem as well. Lastly, the inner product is critical for both the Born rule, which is a transition probability of transition before and after the measurement, and linear decomposition into superposition, which is done at equal time. The goal, then, is to find a characterization of our physical requirement that is interpretation independent, more cohesive and perhaps more physically necessary.

The idea is to reformulate the requirements that lead to superposition and the Born rules in terms of equivalent requirements of statistical mixtures. Mathematically, these will be in terms of the space of density operators in stead of the vector space itself.\footnote{Similar efforts can be found in the literature of Quantum Logic and of Generalized Probability Theories. TODO: add citations}.

The idea is to link superposition to another feature of quantum mechanics: the non-unique decomposition of mixed state into pure states. For example, if we take the maximally mixed state for a spin 1/2 system, this can be implemented with an equal mixture of spin up and spin down, or with an equal mixture of spin left and spin right. The idea is that a pure state $\psi$ is expressible as a superposition of other pure states $\{\phi_i\}$ exactly if there is a mixed state $\rho$ that can be equivalently expressed as a mixture of $\{\phi_i\}$ or as a mixture of $\psi$ and other pure states. To keep the reasoning as general as possible, we will not require $\{\phi_i\}$ to be orthogonal to each other.

\begin{prop}\label{prop_densitySpan}
	Let $\mathcal{H}$ be a complex inner product space. Let $\{\psi_i\}, \{\phi_j\} \in \mathcal{H}$ be two finite sets of normalized vectors. Suppose there exists a density operator that can be expressed as a strict convex combination of either set, that is we can find two sequences $\{a_i\}, \{b_j\} \in \mathbb{R}^+$ such that
	\begin{enumerate}
		\item $\sum_i a_i = 1$
		\item $\sum_j b_j = 1$
		\item $\sum_i a_i |\psi_i\>\<\psi_i| = \sum_j b_j |\phi_j\>\<\phi_j|$,
	\end{enumerate}
	then the span of the two sets is the same.
\end{prop}

@@

\begin{proof}
	The set of all possible density operators is a convex set, which can be understood as a subset of a vector space $V$. Let $\rho$ be a density operator such that $\<\psi|\rho|\psi\>\neq 0$. If $\rho = |\psi\>\<\psi|$, we are done. If not, the two elements will identify a line, which must intersect the boundary of the convex set in two places. One must be $\psi$, since $\psi$ is a pure state and therefore an extreme point. Let $\rho_1$ be the other point. If it is an extreme point, we are done as $\rho$ is an affine combination of $\psi$ and $\rho_1$. If not, the boundary of the convex set in a neighborhood of $rho_1$ is not strictly convex, and therefore is a subspace $V_1$ of the vector space $V$. Note that $V_1$ does not include $\psi$ and it does not span the direction identified by $\rho_1$ and $\psi$. Furthermore, we can't have $\rho_1 = \rho$ as $\<\psi|\rho|\psi\>\neq 0$. Therefore we have that $\rho$ is an affine combination of $\psi$ and $\rho_1$, which must be a density operator of finite rank lower than $\rho$. We can express $\rho_1$ as a convex combination of finitely many extreme points $\psi_2, \psi_3, ..., \psi_n$, each of them corresponding to a pure state. We have that $\rho$ is an affine combination of pure states that include $\psi$.
\end{proof}

\begin{prop}\label{prop_superpositionIsDecomposition}[Superposition is multiple decomposition of mixed states.]
Let $\mathcal{H}$ be a complex vector space. Let $\psi \in \mathcal{H}$ be a normalized vector and $\{\phi_j\} \in \mathcal{H}$ a finte sets of normalized vectors that does not contain $\psi$. Then $\psi$ is a superposition of $\{\phi_j\}$ if and only if there exists a density operator $\rho$ that can be expressed both as a strict convex combination of $\{\phi_i\}$ and of another set of normalized vectors that contain $\psi$. That is, we can find a countable set of normalized vectors $\{\psi_i\} \in \mathcal{H}$ and two sequences $\{a_i\}, \{b_j\} \in \mathbb{R}^+$ such that
\begin{enumerate}
	\item $\psi_1 = \psi$
	\item $\sum_i a_i = 1$
	\item $\sum_j b_j = 1$
	\item $\sum_i a_i |\psi_i\>\<\psi_i| = \sum_j b_j |\phi_j\>\<\phi_j|$,
\end{enumerate}
\end{prop}

\begin{proof}
	Suppose $\psi$ is a normalized state that can be written as $\psi = \sum_{i=1}^n c_i \phi_i$ where $\{\phi_i\} \in \mathcal{H}$ a finte sets of normalized vectors that do not contain $\psi$. Since $\{\phi_j\}$ are not necessarily orthogonal, it is not necessarily true that $|\<\psi|\phi_i\>| = |c_i|^2 \neq 0$. However, since $\psi$ is normalized, there must be at least two vectors in $\{\phi_i\}$ such that $|\<\psi|\phi_i\>| \neq 0$. We can then find $\{p_i\} \in \mathbb{R}+$ for which $\sum_i p_i =1$, such that $\rho = \sum_i p_i |\phi_j\>\<\phi_j|$ and $\<\psi|\rho|\psi\> \neq 0$. By Proposition \ref{prop_decomposition}, $\rho$ can be expressed as a convex combination of a set of normalized states that include $\psi$.

@@

We can now show that the inner product space structure is physical in the sense that it is exactly the structure that is needed to model quantum mixed states and their entropy. One key difference between classical and quantum mechanics is that there is no single decomposition of quantum mixed states in terms of pure states. We can show that relationships of superposition are equivalent to relationships of multiple decomposition.

\begin{proof}
	Let's first address the trivial case in which $\phi_1 = \phi_2$. Suppose that $\psi = c_1 \phi_1 + c_2 \phi_2$, then $\psi = \phi_1$. We set $\bar{\psi} = \psi$ and find $p_1|\psi\>\<\psi| + p_2|\bar{\psi}\>\<\bar{\psi}| = \frac{1}{2}|\phi_1\>\<\phi_1| + \frac{1}{2}|\phi_2\>\<\phi_2|$ for any $p_1, p_2 \in \mathbb{R}$ such that $p_1+p_2=1$. Conversely, suppose $p_1|\psi\>\<\psi| + p_2|\bar{\psi}\>\<\bar{\psi}| = \frac{1}{2}|\phi_1\>\<\phi_1| + \frac{1}{2}|\phi_2\>\<\phi_2|$ for some $p_1, p_2 \in \mathbb{R}$ such that $p_1+p_2=1$. Then $p_1|\psi\>\<\psi| + p_2|\bar{\psi}\>\<\bar{\psi}| = |\phi_1\>\<\phi_1|$ which can be true only if $\psi=\bar{\psi}=\phi_1$.

	Now let's address the case in which $\phi_1 \neq \phi_2$. Suppose that $\psi = c_1 \phi_1 + c_2 \phi_2$. Then $\psi$ is part of the subspace defined by $\phi_1$ and $\phi_2$, which means all the states can be represented as points on the surface of a Bloch sphere. The mixed state $\rho=\frac{1}{2}|\phi_1\>\<\phi_1| + \frac{1}{2}|\phi_2\>\<\phi_2|$ is the midpoint of the chord connecting $\phi_1$ and $\phi_2$. Now, take the line that connects $\rho$ with $\psi$. This line will traverse the sphere in another point $\hat{\psi}$. As $\rho$ sits on the chord between $\psi$ and $\bar{\psi}$, it is a mixture of the two states for some $p_1$ and $p_2$. Conversely, suppose $\rho = p_1|\psi\>\<\psi| + p_2|\bar{\psi}\>\<\bar{\psi}| = \frac{1}{2}|\phi_1\>\<\phi_1| + \frac{1}{2}|\phi_2\>\<\phi_2|$. This means that the subspace spanned by $\psi$ and $\bar{\psi}$ is the same as the one spanned by $\phi_1$ and $\phi_2$, which means that $\psi = c_1 \phi_1 + c_2 \phi_2$ for some $c_1, c_2 \in \mathbb{C}$ such that $|p_1|^2+|p_2|^2=1$.
\end{proof}

As for the inner product, we show that it can be fully characterized, instead of by the Born rule, but the entropy of equal mixtures over state pairs. This gives us a geometric understanding of the inner product at equal time, instead of a probability of transition.

@@

We are then able to argue for the following
\begin{prop}
	The description of quantum states as rays of a complex inner product space is physically justified.
\end{prop}
\begin{justification}
	Since physical preparations and measurements are most of the type not perfect, a physical theory must define statistical ensembles. Part of that definition includes keeping track statistical ensembles under statistical mixtures, that is which particular mixture is obtained by mixing particular states in particular proportions. A particular feature of quantum mixed states is that, unlike classical states, they do not in general provide a unique decomposition in terms of pure states. This non-unique decomposition, as shown by proposition \ref{prop_superpositionIsDecomposition}, is equivalent to the ability to create superpositions of pure state. The inner product structure is therefore the exact structure needed to characterize quantum ensembles and their properties, which is required by the physics.

>>>>

In this appendix we provide the same arguments presented in the main body of the article expanded with the full mathematical detail.

@@

The first task is to show that the vector space structure is not, by itself, physical. Since quantum states are represented by rays of a Hilbert space, and since rays can be defined simply on top of the vector space structure alone, one may think that the vector space structure, by itself, is physical. The issue is that, without an inner product, one cannot make physical sense of the coefficients one uses for linear combinations, therefore superpositions becomes ill-defined.

To show the problem, we prove the following proposition that tells us that there are infinitely many ways to decompose a vector representing a state in terms of vectors that represent two other states, meaning that the coefficients of the linear combination are arbitrary.
\begin{prop}\label{vector_insufficient}[The vector space structure is not enough to uniquely identify superpositions.] Let $\mathcal{H}$ be a complex vector space and let $\textbf{P}(\mathcal{H})$ be its projective space. Let $\psi, \phi_1, \phi_2 \in \mathcal{H}$ such that $\psi = c_1 \phi_1 + c_2 \phi_2$ for some $c_1, c_2 \in \mathbb{C}$. Let $\pj{\psi}, \pj{\phi_1}, \pj{\phi_2} \in \textbf{P}(\mathcal{H})$ be the corresponding rays in the projective space. Then for any $\hat{c}_1, \hat{c}_2 \in \mathbb{C}$ we can find $\hat{\phi}_1 \in \pj{\phi_1}$ and $\hat{\phi}_2 \in \pj{\phi_2}$ such that
$$\psi = \hat{c}_1 \hat{\phi}_1 + \hat{c}_2 \hat{\phi}_2.$$
\end{prop}

@@

The intuition that a state has a single linear decomposition in terms of other vectors requires, first of all, that the vectors are normalized. To identify the coefficient as probability amplitudes, that is that $\sum_i |c_i|^2 = 1$, further requires the decomposition to be done on orthogonal vectors.\footnote{Mathematically, this means turning the vectors $\{\phi_i\}$ into an orthonormal set using Gram-Schmidt orthonormalization, which requires the inner product.} That is, linear combinations require the inner product to be understood physically as superpositions. The vector space structure by itself is therefore insufficient to provide a clear connection to physics.

@@

First of all, it is not clear whether superpositions are actual physical entities or whether they are just mathematical constructs. This is an area where different interpretations may differ,\footnote{TODO: add citations} and the physicality of a mathematical representation should not be interpretation dependent. Second, the Born rule, in its standard characterization, brings in one of the most controversial aspects of the theory, measurements, which suffers from an interpretation problem as well. Lastly, the inner product is critical for both the Born rule, which is a transition probability of transition before and after the measurement, and linear decomposition into superposition, which is done at equal time. The goal, then, is to find a characterization of our physical requirement that is interpretation independent, more cohesive and perhaps more physically necessary.

The idea is to reformulate the requirements that lead to superposition and the Born rules in terms of equivalent requirements of statistical mixtures. Mathematically, these will be in terms of the space of density operators instead of the vector space itself.\footnote{Similar techniques can be found in the literature of Quantum Logic and of Generalized Probability Theories. TODO: add citations}

The idea is to link superposition to another feature of quantum mechanics: the non-unique decomposition of mixed state into pure states. For example, if we take the maximally mixed state for a spin 1/2 system, this can be implemented with an equal mixture of spin up and spin down, or with an equal mixture of spin left and spin right. The idea is that a pure state $\psi$ is expressible as a superposition of other pure states $\{\phi_i\}$ exactly because there is a mixed state $\rho$ that can be equivalently expressed as a mixture of $\{\phi_i\}$ or as a mixture of $\psi$ and other pure states. Hilbert space subspaces are grouping together pure states that can provide the same statistical mixtures. To keep the reasoning as general as possible, we will not require $\{\phi_i\}$ to be orthogonal to each other.

\begin{prop}\label{prop_densitySpan}
	Let $\mathcal{H}$ be a complex inner product space. Let $\{\psi_i\}, \{\phi_j\} \in \mathcal{H}$ be two finite sets of normalized vectors. Suppose there exists a density operator that can be expressed as a strict convex combination of either set, that is we can find two sequences $\{a_i\}, \{b_j\} \in \mathbb{R}_{>0}$ such that
	\begin{enumerate}
		\item $\sum_i a_i = 1$
		\item $\sum_j b_j = 1$
		\item $\sum_i a_i |\psi_i\>\<\psi_i| = \sum_j b_j |\phi_j\>\<\phi_j|$,
	\end{enumerate}
	then the span of the two sets is the same.
\end{prop}

@@

\begin{proof}
	The set of all possible density operators is a convex set, which can be characterized as a subset of a vector space $V$. Let $\rho$ be a density operator such that $\<\psi|\rho|\psi\>\neq 0$. This will be a point in the convex set. If $\rho = |\psi\>\<\psi|$, we are done. If not, the two elements will identify a line, which must intersect the boundary of the convex set in two places. One must be $\psi$, since $\psi$ is a pure state and therefore an extreme point. Let $\rho_1$ be the other point. If it is an extreme point, we are done as $\rho$ is an affine combination of $\psi$ and $\rho_1$. If not, $\rho_1$ is on a part of the boundary that is not strictly convex. That region will be a convex set constrained on a subspace $V_1$ of $V$. Note that $V_1$ does not include $\psi$ and it does not span the direction identified by $\rho_1$ and $\psi$. This means that we can't have $\rho_1 = \rho$ as $\<\psi|\rho|\psi\>\neq 0$. Therefore we have that $\rho$ is an affine combination of $\psi$ and $\rho_1$, which must be a density operator of finite rank lower than $\rho$. We can express $\rho_1$ as a convex combination of finitely many extreme points $\psi_2, \psi_3, ..., \psi_n$, each of them corresponding to a pure state. We have that $\rho$ is an affine combination of pure states that include $\psi$.
\end{proof}

\begin{prop}\label{prop_superpositionIsDecomposition}[Superposition is multiple decomposition of mixed states.]
Let $\mathcal{H}$ be a complex vector space. Let $\psi \in \mathcal{H}$ be a normalized vector and $\{\phi_j\} \in \mathcal{H}$ a finte sets of normalized vectors that does not contain $\psi$. Then $\psi$ is a superposition of $\{\phi_j\}$ if and only if there exists a density operator $\rho$ that can be expressed both as a strict convex combination of $\{\phi_i\}$ and of another set of normalized vectors that contain $\psi$. That is, we can find a finite set of normalized vectors $\{\psi_i\} \in \mathcal{H}$ and two sequences $\{a_i\}, \{b_j\} \in \mathbb{R}_{>0}$ such that
\begin{enumerate}
	\item $\psi_1 = \psi$
	\item $\sum_i a_i = 1$
	\item $\sum_j b_j = 1$
	\item $\sum_i a_i |\psi_i\>\<\psi_i| = \sum_j b_j |\phi_j\>\<\phi_j|$,
\end{enumerate}
\end{prop}

\begin{proof}
	Suppose $\psi$ is a normalized state that can be written as $\psi = \sum_{i=1}^n c_i \phi_i$ where $\{\phi_i\} \in \mathcal{H}$ is a finite sets of normalized vectors that do not contain $\psi$. Since $\{\phi_j\}$ are not necessarily orthogonal, it is not necessarily true that $|\<\psi|\phi_i\>| = |c_i|^2$ or that $|\<\psi|\phi_i\>| \neq 0$ for all $i$. However, since $\psi$ is normalized and $\psi \neq \phi_i$ for all $i$, there must be at least two vectors in $\{\phi_i\}$ such that $|\<\psi|\phi_i\>| \neq 0$. We can then find $\{p_i\} \in \mathbb{R}+$ for which $\sum_i p_i =1$, such that $\rho = \sum_i p_i |\phi_j\>\<\phi_j|$ and $\<\psi|\rho|\psi\> \neq 0$. By Proposition \ref{prop_decomposition}, $\rho$ can be expressed as a convex combination of a set of normalized states that include $\psi$.

@@

As for the inner product, we show that the Born rule can be understood as characterizing the entropy of equal mixtures over state pairs. This gives us a geometric understanding of the inner product at equal time, instead of a probability of transition, which brings it more in line with the use during superposition. It should also be noted that the von Neumann entropy of a mixture can be understood as the minimum Shannon entropy over all possible decompositions,\footnote{TODO: cite} which again shows the connection between the geometry of the convex space of statistical ensembles and the inner product.

@@

Having shown the connection between superposition and non-unique decomposition of mixed state, and between the Born rule and the entropy of mixtures, we can now use these to justify the physicality of the complex inner product space structure for quantum states.
\begin{prop}
	The description of quantum states as rays of a complex inner product space is physically justified.
\end{prop}
\begin{justification}
	Since physical preparations and measurements are most of the type not perfect, a physical theory must define statistical ensembles. Part of that definition includes keeping track statistical ensembles under statistical mixtures, that is which particular mixture is obtained by mixing particular states in particular proportions. A particular feature of quantum mixed states is that, unlike classical states, they do not in general provide a unique decomposition in terms of pure states. This non-unique decomposition, as shown by proposition \ref{prop_superpositionIsDecomposition}, is equivalent to the ability to create superpositions of pure state. The inner product structure is therefore the exact structure needed to characterize quantum ensembles and their properties, which is required by the physics.
